Fast and scalable Gaussian process modeling of stellar variability \vspace{-3\baselineskip}

### 2016-10-22 paper/ms.tex
Added discussion of Gaussian, Matern 5/2

The functional form of $K_j$ may be used to approximate some commonly used Kernel components: radial
kernels can be approximated with $\beta_I=0$, while periodic kernels can be approximated by
$\beta_I \ne 0$.  The squared exponential (Gaussian) kernel and the Mat\'ern kernels are commonly
used radial kernels, while the cosine kernel and the exponential-sine-squared kernels are
commonly used periodic kernels.  Linear combinations of our basis of damped cosine kernels may be 
used to approximate each of these kernels, with various degrees of success:
\begin{itemize}
\item {\bf Constant kernel} A constant kernel component can be used to capture long-timescale
correlations in data, and can be represented with $\beta =  0$.
\item {\bf Exponential kernel.} This is also known as a Mat\'ern 1/2 kernel, and is exactly
represented with $\beta_I=0$.
\item The squared exponential kernel has the behavior of declining steeply at large
time separation, and has $c_1$ = 0.  An infinite number of exponential kernels may be used to
approximate a Gaussian kernel.  For example, to

\item {\bf Mat\'ern 3/2 kernel.}  The Mat\'ern $3/2$ kernel is given by: 
\begin{equation}
K_{3/2}(\tau,\tau_0) = (1+x)e^{-x},
\end{equation} 
where $x = \sqrt{3}\vert \tau/\tau_0\vert$.  This kernel has a Fourier Transform\footnote{Using the normalization convention: 
$\hat f(\omega) = \frac{1}{\sqrt{2\pi}} \int_{-\infty}^\infty f(t) e^{i\omega t} dt$}  of
\begin{equation}
\hat K_{3/2}(\omega) = \frac{\tau_0 6^{3/2}}{\pi^{1/2}} (3+\tau_0^2\omega^2)^{-2},
\end{equation}
where $\omega = 2\pi f$.  This kernel may be well approximated by the sum of two exponential kernels: 
\begin{equation}
K_{3/2}(\tau,\tau_0) \approx (1-b)e^{-x}+be^{-x\frac{b-1}{b}},
\end{equation} 
where $b \gg 1$ is a dimensionless parameter; this approximation is exact in the limit $b \rightarrow \infty$.  For
large values of $b$, the maximum error on this scales as $2/(e^2b) \approx 0.27/b$; for
$b=10$, the error in this approximation is $<3$\%.  Note that the two exponential terms
have opposite signs, which may lead to numerical instability for values of $b$ that
are large.

One advantage of the Mat\'ern 3/2 kernel is that it is smooth across
$\tau=0$ since its derivative is zero at $\tau=0$.  The exponential kernel {\it does not}
have this property, which leads to qualitatively different noise properties on small
timescales due to the sharpness of the exponential kernel at $\tau=0$.  The Mat\'ern 3/2 kernel
turns out to be an excellent description of stellar activity and granulation, and thus the approximate
formula is simultaneously physically relevant and computationally convenient.  
The power spectrum of a single exponential scales as $f^{-2}$ at high frequency,
while the Mat\'ern kernel scales as $f^{-4}$ which more closely matches the power
spectrum due to activity and granulation in the Sun \citep{2004A&A...414.1139A,2009A&A...495..979M}.
%\item Matern 5/2 kernel.
\item {\bf Cosine kernel.} The cosine kernel may be represented exactly with $\beta_R=0$.
\item exp-sin-square kernel
\end{itemize}

>>>>

The functional form of $K_j$ may be used to approximate some commonly used Kernel components: radial
kernels can be approximated with $\beta_I=0$, while periodic kernels can be approximated by
$\beta_I \ne 0$.  The squared exponential (Gaussian) kernel and the Mat\'ern kernels are commonly
used radial kernels, while the cosine kernel and the exponential-sine-squared kernels are
commonly used periodic kernels.  Linear combinations of our basis of damped cosine kernels may be 
used to approximate each of these kernels, with various degrees of success:
\begin{itemize}
\item {\bf Constant kernel} A constant kernel component can be used to capture long-timescale
correlations in data, and can be represented with $\beta =  0$.
\item {\bf Exponential kernel.} This is also known as a Mat\'ern 1/2 kernel, and is exactly
represented with $\beta_I=0$.
\item {\bf Cosine kernel.} The cosine kernel may be represented exactly with $\beta_R=0$.
\item exp-sin-square kernel
\item {\bf Squared exponential.} The squared exponential or Gaussian kernel, $e^{-(\tau/\tau_0)^2/2}$, 
has the behavior of declining steeply at large time separation, is an even function, and has $c_1$ = 0.  
An infinite number of cosine kernels may be used to approximate a Gaussian kernel, but any
finite combination of cosine kernels leads to a negative amplitude at finite $\tau$.  Thus,
in general it is difficult to model a squared exponential with sums of damped cosine terms due
to the heavier tail.  However, the Mat\'ern kernel may be used to approximate it on short
timescales.
\item {\bf Mat\'ern $p+1/2$ kernel.}  The Mat\'ern kernel is given by:
\begin{equation}
C_{p+1/2}(x) = \sigma^2 e^{-\sqrt{(2p+1)}x} \frac{\Gamma(p+1)}{\Gamma(2p+1)}
\sum_{i=0}^p \frac{(p+i)!}{i!(p-i)!} ({\sqrt{(8p+4)}x)^{p-i},
\end{equation} 
where $p$ is a positive integer.  In the limit $p\rightarrow \infty, this becomes the
squared exponential kernel: $C_\infty(x) = \sigma^2 e^{-x^2/2}$.  But, for small
values of $p$, the Mat\'ern kernel may be approximated well by the sums of
exponential kernels, described next.
\item {\bf Mat\'ern 3/2 kernel.}  The Mat\'ern $3/2$ kernel is given by: 
\begin{equation}
K_{3/2}(\tau,\tau_0) = (1+x)e^{-x},
\end{equation} 
where $x = \sqrt{3}\vert \tau/\tau_0\vert$.  This kernel has a Fourier Transform\footnote{Using the normalization convention: 
$\hat f(\omega) = \frac{1}{\sqrt{2\pi}} \int_{-\infty}^\infty f(t) e^{i\omega t} dt$}  of
\begin{equation}
\hat K_{3/2}(\omega) = \frac{\tau_0 6^{3/2}}{\pi^{1/2}} (3+\tau_0^2\omega^2)^{-2},
\end{equation}
where $\omega = 2\pi f$.  This kernel may be well approximated by the sum of two exponential kernels: 
\begin{equation}
K_{3/2}(\tau,\tau_0) \approx (1-b)e^{-x}+be^{-x\frac{b-1}{b}},
\end{equation} 
where $b \gg 1$ is a dimensionless parameter; this approximation is exact in the limit $b \rightarrow \infty$.  For
large values of $b$, the maximum error on this scales as $2/(e^2b) \approx 0.27/b$; for
$b=10$, the error in this approximation is $<3$\%.  Note that the two exponential terms
have opposite signs, which may lead to numerical instability for values of $b$ that
are large.

One advantage of the Mat\'ern 3/2 kernel is that it is smooth across
$\tau=0$ since its derivative is zero at $\tau=0$.  The exponential kernel {\it does not}
have this property, which leads to qualitatively different noise properties on small
timescales due to the sharpness of the exponential kernel at $\tau=0$.  The Mat\'ern 3/2 kernel
turns out to be an excellent description of stellar activity and granulation, and thus the approximate
formula is simultaneously physically relevant and computationally convenient.  
The power spectrum of a single exponential scales as $f^{-2}$ at high frequency,
while the Mat\'ern kernel scales as $f^{-4}$ which more closely matches the power
spectrum due to activity and granulation in the Sun \citep{2004A&A...414.1139A,2009A&A...495..979M}.
\item {\bf Mat\'ern 5/2 kernel.}  The Mat\'ern 5/2 kernel is given
by
\begin{equation}
C_{5/2}(x) = \sigma^2 (1+y+y^2/3)e^{-y},
\end{equation}
where $y = \sqrt{5}x$.  The Fourier transform is:
\begin{equation}
\hat K_{5/2}(\omega) = \frac{2\tau_0 10^{5/2}}{3\pi^{1/2}} (5+\tau_0^2\omega^2)^{-3}.
\end{equation}
This kernel may be approximated by the sum of three exponential
kernels, and has a steeper dependence at large frequencies, $f^{-6}$.  We aren't aware of a stellar
power spectrum that requires such a steep spectrum, and thus we forgo implementation of this kernel.
\end{itemize}
